Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-01 manu-draft.qmd
adding refs to discussion

Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022. Also noted the trend was most pronounced in the winter, see Fig 3.

Preprint about warming SW Fl estuaries: <https://www.researchsquare.com/article/rs-3528678/v1>

@@

@@

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments. At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both crossed thresholds showed consistent increases across the period of record. Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.

Similar regional, long-term changes in coastal waters have been observed by others. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico. Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the authors note that the power of detecting the trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein. @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments. However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area. These differences are understandable given the analysis goals and methods for data synthesis. Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico. These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. Regardless, @nickerson2023 state a Gulf-wide increase in water temperature of 0.1 - 0.5 $^\circ$C per decade, which is comparable to our estimates.

Our relatively simple modeling approach provided some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay. The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drivers. However, a weight-of-evidence approach evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models in explaining noisy ecological patterns. Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period. For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric such that a negative association was observed post-2016 as compared to pre-2016 where frequency occurrence of seagrass declined with the increasing number of days when salinity was below 25 ppt. Likewise, the FIM models had a significant interaction term for the association of the temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016. An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify increases in annual stress associated with temperature and salinity, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling. As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs. These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. These thresholds are not to be considered ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the futre. Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement). Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth. This relates to the inclusion of additional drivers as well as issues related to correlation/causation. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences could additionally be explained by temperature or salinity. Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue. Modeling seagrass change as related to temperature or salinity for the whole record without time period as a predictor would have shown a spurious correlation of both with seagrass given the long-term recovery of Tampa Bay related to nutrient reduction, increased water clarity, and increased seagrass coverage. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen). This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses. This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our weight-of-evidence modeling approach suggest that climate-related stressors driving seagrass loss is a more likely scenario.

The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area [@lizcano-sandoval2022]. Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south. These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay. Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist. Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive. For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth. These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times. The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [@usup1994; @phlips2006; @lopez2021].

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present. Competitive differences between seagrasses and macroalgae are poorly understood in these systems, in addition to insufficient data to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats. In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.

@@

Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at supporting ecological resources in Tampa Bay. Management actions that have historically been effective may have degraded capacity to support ecosystem status due to changing resilience of ecological resources with climate change.  For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay.  As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated.  Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modification).  These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where trends have been most pronounced.  Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded systems due to hysteresis.

@@

<!-- {{< pagebreak >}} -->

>>>>

Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022. Also noted the trend was most pronounced in the winter, see Fig 3

@@

el nino effects, see Greening and Janicki 2006

@@

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments [@osland2015; @oliver2018]. At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, the Bay has demonstrated a consistent trend towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most consistently observed in the EPC dataset with the longest period of record covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in the stressor metrics, where the continuous number of days each year when temperature was above, salinity was below, or both crossed thresholds showed consistent increases across the period of record. Notably, bottom and surface trends for water temperature and salinity in the EPC dataset were similar, suggesting that any potential stress on biotic resources related to these factors are likely occurring throughout the water column.

Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@nickerson2023; @shiir]. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico. Temperature trends were similar for Tampa Bay (the EPC dataset was also used), although the power to detect trends decreased for shorter time periods, as independently verified with the FIM and PDEM datasets herein. @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, which somewhat conflicts with our results showing distinct increases in the summer, particularly for the OTB and HB segments. However, our assessment evaluated trends at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to the larger spatial area. These differences are understandable given the analysis goals and methods for data synthesis. Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico. These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico. The upper limit of our warming estimates for Tampa Bay is comparable.

Our relatively simple modeling approach provided some evidence that climate-related stressors can explain the recent loss in seagrass in Tampa Bay. The models did not provide a consistent, nor statistically powerful, explanation that increasing temperature and decreasing salinity were key drivers. However, evaluating all models together does suggest a pattern that demonstrates the value of considering multiple datasets and models to explain noisy ecological patterns. Model results for the EPC and FIM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss post-2016, described primarily using separate interaction terms of temperature or salinity with time period. For the EPC model, the interaction term was not significant for temperature, whereas the interaction was marginally significant for the salinity metric such that a negative association was observed post-2016 as compared to pre-2016 where frequency occurrence of seagrass declined with the increasing number of days when salinity was below 25 ppt. Likewise, the FIM models had a significant interaction term for the association of the temperature with time period and a marginally significant interaction term for the association of salinity with time period, such that percent cover of seagrasses showed a decline with increasing temperature and decreasing salinity post-2016. An important distinction between the EPC and FIM models is that the former evaluated the number of days above/below thresholds each year to quantify increases in annual stress associated with temperature and salinity, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling. As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring programs. These differences highlight challenges describing noisy relationships in long-term ecological datasets, while also demonstrating the utility of our weight-of-evidence approach to describe these relationships.

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the futre. Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement). Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth. This relates to the inclusion of additional drivers as well as issues related to correlation/causation. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth \[[@greening2014a; @sherwood2017a]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) \[@tbep0696; @Beck20a\]. This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses. This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our models suggest that climate-related stressors driving seagrass loss is a more likely scenario.

The seagrass loss in Tampa Bay since 2016 is a notable phenomenon that is not limited to our study area [@lizcano-sandoval2022]. Losses have been observed throughout southwest Florida during this time period, including Sarasota Bay directly south of Tampa Bay and Charlotte Harbor further south. These regional losses suggest that large-scale stressors are driving these changes, leading us to our initial hypothesis that climate-related stressors could explain the change in Tampa Bay. Based on our results, the losses elsewhere may potentially be explained by temperature and salinity and are worth exploring in other southwest Florida coastal regions and estuaries where long-term datasets exist. Additional factors that could explain these changes are also likely co-occurring with climate stress, some of which are unique to Tampa Bay and others that are likely more pervasive. For Tampa Bay, annual summer/fall blooms of the toxic dinoflagellate *Pyrodinium bahamense* have occurred in OTB since 2008 [@usup1994; @lopez2023] and the specific relationships of these blooms with seagrass change is unclear, although the expectation is that seagrass growth may be limited by the degradation of the light environment with algal growth. These blooms are exacerbated by the hydrologic conditions in OTB that contribute to relatively longer residence times [@phlips2006; @lopez2021]. The effect of warming temperature and decreasing salinity in OTB will further complicate the understanding of how these blooms manifest and persist each year [e.g., @stelling2023].

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@jei2022; @scolaro2023]. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay \[@Beck20b\]. Competitive differences between seagrasses and macroalgae are poorly understood in these systems \[but see @bell1997\], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017]. In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018]. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.

@@

Area managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay. Management actions that have historically been effective may have inadequate capacity to maintain ecosystem status due to changing resilience of ecological resources with climate change. For example, nitrogen load reductions have been effective at restoring Tampa Bay, primarily due to wastewater load allocations considered to be within the nutrient assimilative capacity of Tampa Bay [@greening2006; @greening2014]. As Tampa Bay becomes hotter and fresher, similar load allocations may no longer be effectively assimilated. Strategies can be pursued to reduce allowable load inputs from regulated entities and to create additional resilience through more aggressive management strategies (e.g., additional stormwater controls, hydrological modifications). These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends have been most pronounced. Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring degraded systems due to hysteresis.

@@

This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP). TBEP funding for this work stems from EPA Section 320 Grant Funds, and the TBEP’s local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St. Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget. We are indebted to the hundreds of field and administrative staff for collecting and maintaining the valuable long-term monitoring data used in this study.
